Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2010-09-16 Text/migration.tex
Dick's changes 9/16/10
git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@73 3b7f7548-5c66-0410-83ab-3997167635fc

\begin{abstract}

Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally stable. In one simulation, the planet is inserted prior to the initial eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet interacts with a strong spiral mode, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. {\bf See note in Latex file.}
%The assertion in the preceding sentence needs to be verified by analyzing both runs in more detail -- Am(t), periodograms, images zoomed in on the planet to check for arms. The migration is observed; it is the relationship to mode interactions that needs to be checked.
Planet orbit eccentricity tends to fluctuate rapidly and ranges between about 0.02 to 0.1.
\end{abstract}

@@

{\it Eccentricity.} Figure \ref{fig:e} shows the evolution of $e$ in the planet simulations. In both cases, the planets were inserted with an approximate circular velocity computed by adding the interior disk mass to the stellar point mass. The presence of GI activity in the $t = 10$ planet run immediately increases $e$ to $\sim 0.08$. In the $t = 0$ planet run, the modest initial $e$ decreases during the 1.5 ORP when it is migrating only due to its own wake. However, once there are strong nonlinear interactions with GI modes, $e$ jumps upward. In both runs, it appears that interaction with well-established GI activity leads to eccentricities between 0.02 and 0.1 that vary in a chaotic way between these extremes on orbit period time scales. The magnitude of $e$ is roughly consistent with the ratio of the sound speed to orbital speed of the disk and the modest values for the Mach numbers of the GI spirals \citep{boley2006a,boley2008}.

@@

%Scott and Aaron: I am at a loss for a good punch line to end with. Help. These are just some random remarks.
The inclusion of a 1$M_J$ mass in the equilibrium disk initiates GIs and dramatically effects the character of the GIs. GIs can cause both inward and outward planet migration and can cause rapid migration in the burst phase.
GIs can have a strong effect on the orbital eccentricity of the included planet.
Further work is necessary to adequately analyze these possible mechanisms, but these simulations give a preliminary indication that GI activity strongly affects planet migration.

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{../Figures/planeta.pdf}
\caption{Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The set of red and black curves correspond to the fiducial run and the $t = 0$ planet run, respectively. {\bf The radial migration plot is being used as a placeholder.}}
\label{fig:Am}
\end{figure}

>>>>

\begin{abstract}
%Meaningless Change
Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally stable. In one simulation, the planet is inserted prior to the initial eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet interacts with a strong spiral mode, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. {\bf See note in Latex file.}
%The assertion in the preceding sentence needs to be verified by analyzing both runs in more detail -- Am(t), periodograms, images zoomed in on the planet to check for arms. The migration is observed; it is the relationship to mode interactions that needs to be checked.
Planet orbit eccentricity tends to fluctuate rapidly and ranges between about 0.02 to 0.1.
\end{abstract}

@@

{\it Eccentricity.} Figure \ref{fig:e} shows the evolution of $e$ in the planet simulations. In both cases, the planets were inserted with an approximate circular velocity computed by adding the interior disk mass to the stellar point mass. The presence of GI activity in the $t = 10$ planet run immediately increases $e$ to $\sim 0.08$. In the $t = 0$ planet run, the modest initial $e$ decreases during the 1.5 ORP when it is migrating only due to its own wake. However, once there are strong nonlinear interactions with GI modes, $e$ jumps upward. In both runs, it appears that interaction with well-established GI activity leads to eccentricities between 0.02 and 0.1 that vary in a chaotic way between these extremes on orbit period time scales. The magnitude of $e$ is roughly consistent with the ratio of the sound speed to orbital speed of the disk and the modest values for the Mach numbers of the GI spirals \citep{boley2006a,boley2008}. In other words, the planet's orbit tends to have about as much nonaxisymmetry as reflected in the pitch angles of the GI-induced spiral \citep{cossins2009}.

@@

{\it Summary.} The simulations in this Letter show that a planet of even only 1$M_J$ placed near the $Q$-minimum in a 0.14$M_{\odot}$ disk can hasten the onset of gravitational instabilities and change the nature of the dominant modes during the initial burst. Although the magnitude of radial migration is not too different in magnitude than computed from the simplest isothermal disk formula based on the Lindblad torques (a few$\times10^{-3}$ AU yr$^{-1}$), it can fluctuate in sign on time scales on the order of orbit periods and can be strongly affected by gravitational interactions of the planet with discrete GI-driven spiral modes, leading to phases of rapid radial motion. There is some suggestion in the simulations that planet migration may stall for long periods near the ILRs of dominant global GI-modes. 
%Scott and Aaron: I am at a loss for a good punch line to end with.

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{../Figures/planeta.pdf}
\caption{Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The set of red and black curves correspond to the fiducial run (no planet) and the $t = 0$ planet run, respectively. {\bf The radial migration plot is being used as a placeholder.}}
\label{fig:Am}
\end{figure}
